Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-04-10 VegOzone.tex
Checked with Grammarly

In our assessment of primary drivers of ozone damage risk on vegetation in the subarctic, we focus on the 2018 growing season which was subject to an extended heatwave and its subsequent growing season of 2019. The NIBIO operated Environment Centre Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley serves as main reference. Throughout the paper, we will use Svanhovd synonymously for the atmospheric monitoring site at Svanvik/Pasvik. The available meteorological monitoring data are described in Section~\ref{subsec:atmo_svanvik}. Other data used for cross-evaluation of the ozone monitoring data taken at Svanhovd is described in Section~\ref{subsec:other_data}. We compare with observational data from other monitoring sites in northern Fennoscandia (Section~\ref{subsubsec:ebas}) and use these to evaluate reanalysis products for ground-level ozone (Section~\ref{subsubsec:ozone_rea}).

\subsection{NIBIO Environment Centre Svanhovd and atmospheric monitoring}
\label{subsec:atmo_svanvik}
The foundation of the NIBIO Environment Centre Svanhovd dates back to the 1930s \citep{NIBIO_Svanhovd}. The center focuses on research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.e. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths. The locations of the atmospheric monitoring site and the ozone garden are marked in an aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a). Svanhovd is located south of the settlement of Svanvik and the border to Russia, the Pasvik river, lies only a few hundred meters to the east.\\

@@

An ozone garden was installed in the course of our project at the NIBIO Environment Centre Svanhovd and was likely the northern-most ozone garden to date. An ozone garden consists of selected plant species which are more sensitive to ozone and display visible injuris. Cultivated species were, e.g. clover, tobacco, and potatoes. As shown in Fig.~\ref{fig:svanhovd_research_station}b), the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

For the purpose of relating observed damage on ozone sensitive plants to local ozone concentrations, an ozone monitor had been installed during the growing seasons 2018/19 in addition to existing monitors of atmospheric key variables. All variables were obtained through \href{luftkvalitet.no}{luftkvalitet.no} operated by \citet{NILU_AIRQ}. Our ozone monitoring data has also been added to EBAS database operated by \citet{NILU_EBAS} and is thus openly accessable.
Long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.
All relevant data for the growing season 2018/19 are shown in Fig.~\ref{fig:data_svanvik_2018_2019}. Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. The hatched areas mark times when no ozone data was recorded. Note, while the downtime during winter was planned, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

 As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, since data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden which occurred from July 12--29 \citep{SOU2019}. However, due to the above mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak ozone concentration. In contrast, ozone concentration only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.
Hourly averaged $2\,\unit{m}$ temperatures above $20\,\unit{^\circ C}$ occurred more regularly in July 2018 than in 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}b). In 2018, spring temperature regularly rose above freezing only in May, while in 2019 this occurred already early in March/April.
More rain events with accumulated daily precipitation ($\sum_d \mathrm{Precip}$) above $10\,\unit{mm}$ occurred in the summer of 2018 compared to 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}c).
Qualitatively, global irradiance $Q_0$ (Fig.~\ref{fig:data_svanvik_2018_2019}d) was higher in spring (May) and summer (July) 2018 compared to 2019, while June 2019 showed higher irradiance than 2018. A high global irradiance in most cases is the results of a low cloud fraction. In both years, the maximum recorded $Q_0$ was roughly $750\,\unit{W\,m^{-2}}$ reached in June as naturally expected.

@@

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the ozone conditions in 2018, we derive a ground-level ozone concentration climatology for Svanvik. From 1986--1996 ozone monitoring has been operated continously by NILU at Svanhovd, but no ozone measurements have been conducted until the start of our project. Due to probable changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors which fell under air quality regulations in Europe, we will cross-calibrate our data with long-term ozone observations at other sites in northern Fennoscandia. In regard to ozone, available observational data are sparse in this region. An overview over the geographical locations of conducted ozone monitoring in the past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data are described in Section~\ref{subsubsec:ebas}. 
As additional complement to observational data which are not available at all times and spatially sparse, we also evaluate various ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{station_map_fennoscandia}
  \includegraphics[width=8.3cm]{fig03}
  \caption{Cap of the North. Locations of past and present ozone observation sites in northern Fennoscandia used within this study. For more details see Tab.~\ref{tab:ebas_obs}. The same color coding for each station as displayed herein will be used in all following figures.}
  \label{fig:station_map_fennoscandia}
\end{figure}

\subsubsection{Ground-level ozone observations}
\label{subsubsec:ebas}
All ground-level ozone observations used in this study are available from the EBAS database operated by \citet{NILU_EBAS}. In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long-term observations are located on higher ground than Svanhovd which will result in higher ground-level ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, at close vicinity the vegetation varies substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above tree line (pine forest) and mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass (c.c. Fig.\ref{fig:svanhovd_research_station}).

Data from before the 1990s are available solely from Svanvik and Karasjok. These data, however, do not follow the high quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Especially, ozone monitors did not undergo a regular re-calibration. This probably leads to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a breaking point in the combined data series, denoted as Jergul/Karasjok in the following, which would be of a problem if this was the sole data to compare with. Only few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration, but are shown for completion. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas indicate time periods with insufficient quality control as mentioned above.

@@

\begin{figure*}[t]
  %\includegraphics[width=12cm]{ozone_timeseries_fenoscandic_obs.png}
  \includegraphics[width=12cm]{fig04}
  \caption{Time series of ozone observations in northern Fennoscandia (Tab.~\ref{tab:ebas_obs}) (1986--2019). Data available from EBAS. The hatched areas indicate time periods with insufficient quality control according to \citet{NILU2003}.}
  \label{fig:ozone_timesseries_fenoscandic_obs}
\end{figure*}

@@

Details of reanalysis data sets referred to in this work are listed in Table~\ref{tab:ozone_rea}.
There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone, the Monitoring Atmospheric Composition and Climate (MACC) and the newer Copernicus Atmosphere Monitoring Service reanalysis (CAMSRA). From Copernicus member institutions a CAMS Regional Air Quality (CAMSRAQ) reanalysis ensemble is available for Europe. The CAMSRA and CAMSRAQ are available in near real time, but only the former covers a period of sufficient length for climate analysis (2003 -- present). All reanalyses use different but, at that time, latest versions of the operational weather forecast system (OpenIFS) from ECMWF. They differ substantially in the assimilated observational ozone data. The MACC reanalysis assimilates only satellite derived tropospheric column ozone, while CAMSRA also includes ozone profiles from satellite retrievals. Most importantly for our study, in situ observations from ozone near-surface station networks are only assimilated in the CAMSRAQ reanalysis ensemble. The quality with respect to observations in northern Fennoscandia will be discussed in Section~\ref{subsubsec:clim_ozone}. In summary, the results prevented us from using any of the global reanalysis data.

\begin{table}[t]
  \caption{Global/refional ozone reanalysis products used in this study.}
  \label{tab:ozone_rea}
\begin{tabular}{llcclcc}
\tophline
Name & Provider & \multicolumn{2}{c}{Resolution} & Time period & Meteorological forcing & \chem{O_3} assimilation\\
&              & spatial & temporal & & \\
\middlehline
MACC & ECMWF & $0.75^\circ \times 0.75^\circ$ & 3-hourly & 2003 -- 2012 & OPS & satellite $^\triangledown$\\
CAMSRA & ECMWF & $0.75^\circ \times 0.75^\circ$ & 3-hourly & 2003 -- 2012 $^\dagger$ & ERA5 / OPS $^\ddagger$ & satellite $^\blacktriangledown$\\
CAMSRAQ & Copernicus & $0.1^\circ \times 0.1^\circ$ & 1-hourly & 2014 -- 2018 $^\dagger$ & OPS $^\star$ & in situ $^\vartriangle$\\
\bottomhline
\end{tabular}
\belowtable{$^\dagger$ Subset of the data used in this study. The full data set extends into near real time; $^\ddagger$ ERA5 (2003--2016), OPS (later); $^\star$ EURAD uses WRF for downscaling of operational IFS; $^\triangledown$ MLS, OMI - tropospheric column; $^\blacktriangledown$ SCIAMACHY, MIPAS, MLS, OMI, GOME2, SBUV2 - tropospheric column + profile; $^\vartriangle$ METEO France NRT.} % Table Footnotes
\end{table}

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
The location of Svanvik at $69.45\,\unit{^\circ N}$ suggests a subarctic climate. The derived climate diagram following Walter and Lieth shown in Fig.~\ref{fig:climatediagram} supports this very well. The diagram is based on climatological data from Svanvik (1992--2012) downloaded from LandbruksMeteorologiske Tjeneste. Monthly averaged $2\,\unit{m}$ temperatures (red line) are displayed with standard error of mean (SE) as error band. We chose SE in this case due to the large interannual variability in temperature reflected in the standard deviation. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consecutive months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppens climate classification of a regular subarctic climate (Dfc) \citep{SD:Beck2018}. The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. In the time period 1992--2012, the coldest measured temperature was $-45.2\,\unit{^\circ C}$ observed January 27, 1999, while the highest temperature on July 16 the same year was $29.4\,\unit{^\circ C}$.

@@

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{climatediagram}
  \includegraphics[width=8.3cm]{fig05}
  \caption{Climate diagram following Walter and Lieth. The diagram is based on climatological data for Svanvik/Pasvik (1992--2012) downloaded from LandbruksMeteorologiske Tjeneste. Monthly averaged temperatures (red line) are displayed with standard error of mean as error band. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star.}
  \label{fig:climatediagram}
\end{figure}

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
In the following, we will look at derived ozone climatologies from different perspectives.
The multi-annual means of daily mean ozone, for simplicity here without standard deviations, at observation sites in northern Fennoscandia (refer to Table~\ref{tab:ebas_obs} for details) are shown in Supplement Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}. All climatologies except for Svanvik show peak values of about $48\,\unit{ppb}$ in late April. The spring peak values at Svanvik are lower because of the much lower altitude compared to the other monitoring sites. The annual average ozone concentration \chem{\left<[O_3]\right>} declines throughout the growing season until it reaches its minimum in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. At Svanvik, \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than on sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN). The large spread in case of Svanvik and Janiskoski is due to the lower number of observations.

@@

In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, the two dimensional density distributions for the combined northern Fennoscandia data (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) are shown. Darker colors indicate higher probability to observe these values. On top of the density distributions, $10\,\unit{days}$ averages of daily mean \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} are displayed together with $1 \sigma$ uncertainties and SE, respectively. Splines have been fitted through the data for easier inter-comparison with reanalysis data.
From the northern Fennoscandic climatology, one can deduce several features. The build-up of the ozone spring peak takes about $7.5\,\unit{months}$ from minimum to maximum. A sharp increase starts one month ahead of the estimated begin of greening $G_\mathrm{begin}$ (see Section~\ref{subsec:do3se_parameters} for details) and is in coincidence with rising temperatures and global irradiance. We find day of the year (\unit{doy}) $\left<G_\mathrm{begin}\right> = (150 \pm 14)$ for Svanvik (Supplement Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May. This finding is in line with more extensive studies \citep[][e.g.,]{IJB:Karlsen2007,AFM:Linderholm2006}. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} is reached in late April ($100-120\,\unit{doy}$). The depletion of ozone from maximum to minimum takes only $4.5\,\unit{months}$. The minimum occurs at $25\,\unit{ppb}$ in the beginning of September ($240-250\,\unit{doy}$).

The density distribution for Svanvik and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} display a similar pattern as the northern Fennoscandic climatology (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b). The spring peak apparently occurs slightly earlier (in the beginning of April). As Svanvik is located at lower altitude than the other sites, the earlier decline of the ozone peak may point to a stronger influence of vegetation on the removal of ozone. \chem{CO_2} uptake in coniferous trees can occur as early as \unit{doy} 100 \citep{TB:Kolari2007, TP:Wallin2013}. The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during night time.

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{ozone_climatology_fenoscandic_obs_spline}
  \includegraphics[width=8.3cm]{fig06}
  \caption{Ozone climatologies. Fennoscandia combines all available data from Jergul/Karasjok, Esrange, and Pallas. The densitiy distributions are shown together with a multi-annual mean of daily ozone concentrations \chem{[O_3]} (climatology). Splines have been fitted though daily mean \chem{[O_3]}. The climatology of daily mean ozone is shown together with $1\,\sigma$ standard deviation and standard error of the mean. Since \chem{[O_3]} are strongly correlated for sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN) (see Fig.~\ref{fig:density_distribution}), data of these have been averaged together to derive a climatology for northern Fennoscandia. (a) Northern Fennoscandia; (b) Svanvik.}
  \label{fig:ozone_climatology_fenoscandic_obs_spline}
\end{figure}

>>>>

In our assessment of primary drivers of ozone damage risk on vegetation in the subarctic, we focus on the 2018 growing season which was subject to an extended heatwave and its subsequent growing season of 2019. The NIBIO operated Environment Centre Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley serves as the main reference. Throughout the paper, we will use Svanhovd synonymously for the atmospheric monitoring site at Svanvik/Pasvik. The available meteorological monitoring data are described in Section~\ref{subsec:atmo_svanvik}. Other data used for cross-evaluation of the ozone monitoring data taken at Svanhovd is described in Section~\ref{subsec:other_data}. We compare with observational data from other monitoring sites in northern Fennoscandia (Section~\ref{subsubsec:ebas}) and use these to evaluate reanalysis products for ground-level ozone (Section~\ref{subsubsec:ozone_rea}).

\subsection{NIBIO Environment Centre Svanhovd and atmospheric monitoring}
\label{subsec:atmo_svanvik}
The foundation of the NIBIO Environment Centre Svanhovd dates back to the 1930s \citep{NIBIO_Svanhovd}. The center focuses on the research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.e. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths. The locations of the atmospheric monitoring site and the ozone garden are marked in aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a). Svanhovd is located south of the settlement of Svanvik and the border to Russia, the Pasvik river, lies only a few hundred meters to the east.\\

@@

An ozone garden was installed in the course of our project at the NIBIO Environment Centre Svanhovd and was likely the northern-most ozone garden to date. An ozone garden consists of selected plant species that are more sensitive to ozone and display visible injuries. Cultivated species were, e.g. clover, tobacco, and potatoes. As shown in Fig.~\ref{fig:svanhovd_research_station}b), the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

To relate observed damage on ozone sensitive plants to local ozone concentrations, an ozone monitor had been installed during the growing seasons 2018/19 in addition to existing monitors of atmospheric key variables. All variables were obtained through \href{luftkvalitet.no}{luftkvalitet.no} operated by \citet{NILU_AIRQ}. Our ozone monitoring data has also been added to the EBAS database operated by \citet{NILU_EBAS} and is thus openly accessible.
Long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.
All relevant data for the growing season 2018/19 are shown in Fig.~\ref{fig:data_svanvik_2018_2019}. Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. The hatched areas mark times when no ozone data were recorded. Note, while the downtime during winter was planned, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

 As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, since data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden which occurred from July 12--29 \citep{SOU2019}. However, due to the above-mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak ozone concentration. In contrast, ozone concentration only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.
Hourly averaged $2\,\unit{m}$ temperatures above $20\,\unit{^\circ C}$ occurred more regularly in July 2018 than in 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}b). In 2018, spring temperature regularly rose above freezing only in May, while in 2019 this occurred already early in March/April.
More rain events with accumulated daily precipitation ($\sum_d \mathrm{Precip}$) above $10\,\unit{mm}$ occurred in the summer of 2018 compared to 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}c).
Qualitatively, global irradiance $Q_0$ (Fig.~\ref{fig:data_svanvik_2018_2019}d) was higher in spring (May) and summer (July) 2018 compared to 2019, while June 2019 showed higher irradiance than 2018. A high global irradiance in most cases is the result of a low cloud fraction. In both years, the maximum recorded $Q_0$ was roughly $750\,\unit{W\,m^{-2}}$ reached in June as naturally expected.

@@

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the ozone conditions in 2018, we derive a ground-level ozone concentration climatology for Svanvik. From 1986--1996 ozone monitoring has been operated continuously by NILU at Svanhovd, but no ozone measurements have been conducted until the start of our project. Due to probable changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors which fell under air quality regulations in Europe, we will cross-calibrate our data with long-term ozone observations at other sites in northern Fennoscandia. Concerning ozone, available observational data are sparse in this region. An overview of the geographical locations of conducted ozone monitoring in the past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data are described in Section~\ref{subsubsec:ebas}. 
As an additional complement to observational data which are not available at all times and spatially sparse, we also evaluate various ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{station_map_fennoscandia}
  \includegraphics[width=8.3cm]{fig03}
  \caption{Cap of the North. Locations of past and present ozone observation sites in northern Fennoscandia were used in this study. For more details see Tab.~\ref{tab:ebas_obs}. The same color coding for each station as displayed herein will be used in all the following figures.}
  \label{fig:station_map_fennoscandia}
\end{figure}

\subsubsection{Ground-level ozone observations}
\label{subsubsec:ebas}
All ground-level ozone observations used in this study are available from the EBAS database operated by \citet{NILU_EBAS}. In Table~\ref{tab:ebas_obs}, we give an overview of the exact locations and time range of operational data taking. All stations with available long-term observations are located on higher ground than Svanhovd which will result in higher ground-level ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, in close vicinity, the vegetation varies substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above treeline (pine forest) and mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass (c.c. Fig.\ref{fig:svanhovd_research_station}).

Data from before the 1990s are available solely from Svanvik and Karasjok. These data, however, do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Especially, ozone monitors did not undergo regular re-calibration. This probably leads to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream of the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a breaking point in the combined data series, denoted as Jergul/Karasjok in the following, which would be a problem if this was the sole data to compare with. Only a few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration, but are shown for completeness. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas indicate periods with insufficient quality control as mentioned above.

@@

\begin{figure*}[t]
  %\includegraphics[width=12cm]{ozone_timeseries_fenoscandic_obs.png}
  \includegraphics[width=12cm]{fig04}
  \caption{Time series of ozone observations in northern Fennoscandia (Tab.~\ref{tab:ebas_obs}) (1986--2019). Data are available from EBAS. The hatched areas indicate periods with insufficient quality control according to \citet{NILU2003}.}
  \label{fig:ozone_timesseries_fenoscandic_obs}
\end{figure*}

@@

Details of reanalysis data sets referred to in this work are listed in Table~\ref{tab:ozone_rea}.
There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone, the Monitoring Atmospheric Composition and Climate (MACC) and the newer Copernicus Atmosphere Monitoring Service reanalysis (CAMSRA). From Copernicus member institutions a CAMS Regional Air Quality (CAMSRAQ) reanalysis ensemble is available for Europe. The CAMSRA and CAMSRAQ are available in near real time, but only the former covers a period of sufficient length for climate analysis (2003 -- present). All reanalyses use different but, at that time, the latest versions of the operational weather forecast system (OpenIFS) from ECMWF. They differ substantially in the assimilated observational ozone data. The MACC reanalysis assimilates only satellite-derived tropospheric column ozone, while CAMSRA also includes ozone profiles from satellite retrievals. Most importantly for our study, in situ observations from ozone near-surface station networks are only assimilated in the CAMSRAQ reanalysis ensemble. The quality concerning observations in northern Fennoscandia will be discussed in Section~\ref{subsubsec:clim_ozone}. In summary, the results prevented us from using any of the global reanalysis data.

\begin{table}[t]
  \caption{Global/refional ozone reanalysis products used in this study.}
  \label{tab:ozone_rea}
\begin{tabular}{llcclcc}
\tophline
Name & Provider & \multicolumn{2}{c}{Resolution} & Time period & Meteorological forcing & \chem{O_3} assimilation\\
&              & spatial & temporal & & \\
\middlehline
MACC & ECMWF & $0.75^\circ \times 0.75^\circ$ & 3-hourly & 2003 -- 2012 & OPS & satellite $^\triangledown$\\
CAMSRA & ECMWF & $0.75^\circ \times 0.75^\circ$ & 3-hourly & 2003 -- 2012 $^\dagger$ & ERA5 / OPS $^\ddagger$ & satellite $^\blacktriangledown$\\
CAMSRAQ & Copernicus & $0.1^\circ \times 0.1^\circ$ & 1-hourly & 2014 -- 2018 $^\dagger$ & OPS $^\star$ & in situ $^\vartriangle$\\
\bottomhline
\end{tabular}
\belowtable{$^\dagger$ Subset of the data used in this study. The full data set extends into near real-time; $^\ddagger$ ERA5 (2003--2016), OPS (later); $^\star$ EURAD uses WRF for downscaling of operational IFS; $^\triangledown$ MLS, OMI - tropospheric column; $^\blacktriangledown$ SCIAMACHY, MIPAS, MLS, OMI, GOME2, SBUV2 - tropospheric column + profile; $^\vartriangle$ METEO France NRT.} % Table Footnotes
\end{table}

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
The location of Svanvik at $69.45\,\unit{^\circ N}$ suggests a subarctic climate. The derived climate diagram following Walter and Lieth which is shown in Fig.~\ref{fig:climatediagram} supports this very well. The diagram is based on climatological data from Svanvik (1992--2012) downloaded from LandbruksMeteorologiske Tjeneste. Monthly averaged $2\,\unit{m}$ temperatures (red line) are displayed with a standard error of the mean (SE) as error band. We chose SE in this case due to the large interannual variability in temperature reflected in the standard deviation. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consecutive months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppens climate classification of a regular subarctic climate (Dfc) \citep{SD:Beck2018}. The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. In the period 1992--2012, the coldest measured temperature was $-45.2\,\unit{^\circ C}$ observed January 27, 1999, while the highest temperature on July 16 the same year was $29.4\,\unit{^\circ C}$.

@@

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{climatediagram}
  \includegraphics[width=8.3cm]{fig05}
  \caption{Climate diagram following Walter and Lieth. The diagram is based on climatological data for Svanvik/Pasvik (1992--2012) downloaded from LandbruksMeteorologiske Tjeneste. Monthly averaged temperatures (red line) are displayed with standard error of the mean as error band. Averaged monthly accumulated precipitation (blue bars) is shown with a standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star.}
  \label{fig:climatediagram}
\end{figure}

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
In the following, we will look at derived ozone climatologies from different perspectives.
The multi-annual means of daily mean ozone, for simplicity here without standard deviations, at observation sites in northern Fennoscandia (refer to Table~\ref{tab:ebas_obs} for details) are shown in Supplement Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}. All climatologies except for Svanvik show peak values of about $48\,\unit{ppb}$ in late April. The spring peak values at Svanvik are lower because of the much lower altitude compared to the other monitoring sites. The annual average ozone concentration \chem{\left<[O_3]\right>} declines throughout the growing season until it reaches its minimum in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. At Svanvik, \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than on sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN). The large spread in the case of Svanvik and Janiskoski is due to the lower number of observations.

@@

In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, the two dimensional density distributions for the combined northern Fennoscandia data (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) are shown. Darker colors indicate higher probability to observe these values. On top of the density distributions, $10\,\unit{days}$ averages of daily mean \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} are displayed together with $1 \sigma$ uncertainties and SE, respectively. Splines have been fitted through the data for easier inter-comparison with reanalysis data.
From the northern Fennoscandia climatology, one can deduce several features. The build-up of the ozone spring peak takes about $7.5\,\unit{months}$ from minimum to maximum. A sharp increase starts one month ahead of the estimated begin of greening $G_\mathrm{begin}$ (see Section~\ref{subsec:do3se_parameters} for details) and is in coincidence with rising temperatures and global irradiance. We find day of the year (\unit{doy}) $\left<G_\mathrm{begin}\right> = (150 \pm 14)$ for Svanvik (Supplement Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May. This finding is in line with more extensive studies \citep[][e.g.,]{IJB:Karlsen2007,AFM:Linderholm2006}. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} is reached in late April ($100-120\,\unit{doy}$). The depletion of ozone from maximum to minimum takes only $4.5\,\unit{months}$. The minimum occurs at $25\,\unit{ppb}$ in the beginning of September ($240-250\,\unit{doy}$).

The density distribution for Svanvik and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} display a similar pattern as the northern Fennoscandia climatology (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b). The spring peak occurs slightly earlier (at the beginning of April). As Svanvik is located at a lower altitude than the other sites, the earlier decline of the ozone peak may point to a stronger influence of vegetation on the removal of ozone. \chem{CO_2} uptake in coniferous trees can occur as early as \unit{doy} 100 \citep{TB:Kolari2007, TP:Wallin2013}. The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during nighttime.

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{ozone_climatology_fenoscandic_obs_spline}
  \includegraphics[width=8.3cm]{fig06}
  \caption{Ozone climatologies. Fennoscandia combines all available data from Jergul/Karasjok, Esrange, and Pallas. The density distributions are shown together with a multi-annual mean of daily ozone concentrations \chem{[O_3]} (climatology). Splines have been fitted through daily mean \chem{[O_3]}. The climatology of daily mean ozone is shown together with $1\,\sigma$ standard deviation and standard error of the mean. Since \chem{[O_3]} are strongly correlated for sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN) (see Fig.~\ref{fig:density_distribution}), data of these have been averaged together to derive a climatology for northern Fennoscandia. (a) Northern Fennoscandia; (b) Svanvik.}
  \label{fig:ozone_climatology_fenoscandic_obs_spline}
\end{figure}
